\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{verbatim}
\usepackage{natbib}
\usepackage[masterthesis]{systems-cover}
\usepackage{subfiles}
\usepackage{listings, listings-rust, listings-clojure, listings-json, listings-datalog}

\covernum{ 263-0800-00L	}
\covertitle{ Optimising Distributed Dataflows in Interactive Environments }
\coverauthor{ Nikolas Göbel }
\coversupervisedby{ Frank McSherry, Gustavo Alonso }
\coverdate{ November 2018 - April 2019 }

\begin{document}

\tableofcontents

\newpage

\section{Abstract} \label{abstract}

The capability to maintain sophisticated relational queries over
potentially unbounded data streams is a critical competitive advantage
for data-driven organizations. Traditional approaches to relational
query optimization assume ad-hoc execution, benevolent distribution of
data, and the availability of accurate statistics at planning
time. Established practice in low-latency stream processing assumes
comparatively simpler computational models and relaxed consistency
guarantees.

\emph{Differential computation} generalizes incremental computation to
support data-parallel relational operators and arbitrarily nested,
least fixed point iteration. \emph{3DF} utilizes differential
computation to provide continuous, incremental evaluation of Datalog
queries over data streams. In this work we identify the challenges to
providing these new capabilities to dynamic multi-user environments
and describe solutions adapted from the database and stream processing
communities.

We implement our findings in 3DF and \emph{Differential Dataflow}, an
implementation of differential computation. We evaluate the resulting
system and showcase in particular (I) the use of delta-queries to
maintain joins of arbitrary arity with a constant memory footprint,
(II) the use of worst-case optimal join algorithms to provide
predictable query performance on adversarially skewed data, and (III)
strategies to share resources between clients without giving up
individual optimization opportunities.

\newpage

\section*{Acknowledgements}

I would like to thank Professor Gustavo Alonso for introducing me to
Timely and Differential Dataflow, and for supporting my work on these
systems.

I am beyond grateful to Dr. Frank McSherry for condensing a decade of
learning into mere months, and for expanding my views on system
design.

I would also like to thank Andrea Lattuada for his support, and in
particular for his work on monitoring Timely computations.

\newpage

\section{Introduction and Motivation} \label{intro}
\subfile{sections/intro.tex}
\newpage

\section{Problem Description and Outline} \label{problem}
\subfile{sections/problem.tex}
\newpage

\section{Background} \label{background}
\subfile{sections/background.tex}
\newpage

\section{Design and Implementation of 3DF} \label{3df}
\subfile{sections/3df.tex}
\newpage

\section{Known Techniques and Related Work} \label{known-techniques}
\subfile{sections/known-techniques.tex}
\newpage

\subfile{sections/catalog.tex}

\section{General Implementation Details} \label{implementation}
\subfile{sections/implementation.tex}
\newpage

\section{A Worst-Case Optimal Join Operator} \label{impl-hector}
\subfile{sections/hector-implementation.tex}
\newpage

\section{Other Findings and Future Work} \label{future-work}
\subfile{sections/future-work.tex}
\newpage

\section{Conclusions} \label{conclusions}

We have investigated the problem of optimizing the continuous
maintenance of relational queries for dynamic, multi-user
environments. Besides the basic need for consistent, correct results,
we identified robustness, support for many concurrent users with
general-purpose use cases, and low-latency execution at
high-throughputs as the driving goals behind this work. We thus were
looking for techniques that could bring the domain of analytical, and
graph databases closer to that of high-performance stream processing.

In closing, we want to summarize our findings and how they lead to a
system that satisfies these properties.

\begin{enumerate}
\item
  The cyclic dataflow model in combination with progress tracking of
  partially-ordered logical times, as implemented by the Timely
  Dataflow framework, enables consistent, data-parallel computations
  that exploit fine-grained opportunities for concurrent
  execution. Support for local iteration of regions of the dataflow
  dramatically increase the class of algorithms that can be expressed
  within this model.

\item
  Distributed incremental computation with iteration and support for
  retractions, as implemented by the Differential Dataflow framework,
  makes it possible to efficiently maintain complex relational views
  over data streams.

\item
  Compact, shared representations of historical inputs, as implemented
  by Differential's arrangements, allow the dynamic creation of new
  dataflows with low-latency.

\item
  Declarative, relational languages such as Datalog, in combination
  with an attribute-oriented data model, are expressive enough to
  support a wide range of use cases, and are easily extended to new
  domains, without fundamentally affecting the underlying runtime.

\item
  Deferred synthesis of query plans is essential in dynamic
  environments, where queries might be defined across multiple
  declarative interactions, and where partially constrained plans
  significantly impact the overall system performance.

\item
  Exploiting the linearity of the relational join operator via delta
  queries makes it possible to implement joins of arbitrary arity with
  a constant memory footprint and without significantly impacting
  latencies. Delta queries can be combined to form join operators that
  update incrementally in response to arbitrary input changes.

\item
  Worst-case optimal dataflow join algorithms are compatible with
  delta queries and enable robust, predictable performance
  irrespective of clause order. The performance of the resulting
  operator preserves near real-time performance characteristics, is
  competitive with traditional join-at-a-time approaches on best- and
  average-case queries, and offers asymptotic improvements for a class
  of relevant, cyclic queries. Crucially, these properties are
  preserved on long-running dataflows, even as the underlying data
  distributions change, without the need for synchronized
  re-optimization.

\item
  Extensions beyond purely conjunctive queries can be implemented
  within the same worst-case optimal join framework. We gave
  implementations of binary predicates, stratified negation, and
  constant bindings. Combined with Differential's support for
  fixed-point iteration, these cover the needs of many
  functional-relational query languages. Similarly, we outlined an
  extension to unary relations, that allows remediating categorical
  data as a common source of skew.

\item
  Modeling client parameters via additional joins on parameter
  relations allows both the dynamic re-configuration of existing
  dataflows, as well as the multiplexing of multiple client's
  interests onto a single dataflow. The resources of that dataflow can
  then be shared without limiting optimization opportunities. More
  generally applicable re-use is still possible via arrangements and
  scales significantly better than creating individual dataflows for
  every client.

\end{enumerate}

We identified areas for future work, in particular on automated
policies and hybrid planners, that are required to tie these
primitives into fully-formed systems. Nevertheless we believe that our
findings and their implementation within 3DF provide strong
foundations on which a new class of systems, combining the analytical
power of databases with the performance characteristics of stream
processors, can build.

\newpage

\bibliographystyle{te}
\bibliography{references}

\end{document}
